Optimizing Finite Element Integration through Automated Expression Rewriting and Code Specialization

### 2014-07-31 TOMS_Paper/assembly.tex
Start code specializer section

\begin{figure}
\scriptsize
\lstinputlisting{listings/simple.code}
\label{fig:original-code}\caption{Original code}
\end{figure}

@@

\begin{figure}
\scriptsize
\lstinputlisting{listings/factorized.code}
\label{fig:factorized-code}\caption{Factorized code}
\end{figure}

@@

\begin{figure}
\scriptsize
\lstinputlisting{listings/toexpand.code}
\label{fig:toexpand-code}\caption{Expandable code}
\end{figure}

@@

@@

\subsection{Avoiding Iteration on Zero-blocks with Symbolic Execution}
\label{sec:zeros}

\section{Code Specialization}
\label{sec:code-spec}

\subsection{Standard Compiler Transformations}

\subsection{Precomputation of Invariant Terms}

@@

\subsection{Model-driven Autotuning}

%\begin{algorithm}[t]
%\SetAlgorithmName{LISTING}{}
%\footnotesize
%\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
%~~// K, det = Compute Jacobian (coords) \\
%~~\\
%~~\KwSty{static const double} W[3] ALIGN = $\lbrace$...$\rbrace$\\
%~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
%~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
%~~\\
%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
%~~~~double LI$\_$0[4] ALIGN;\\
%~~~~double LI$\_$1[4] ALIGN;\\
%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
%~~~~$\rbrace$\\
%~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
%~~~~~~\KwSty{$\#$pragma vector aligned}\\
%~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
%~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+LI$\_$1[k]*LI$\_$1[j])*det*W[i]);\\
%~~$\rbrace$\\
%$\rbrace$
%\caption{Local assembly code for the Helmholtz problem in
%  Listing~\ref{code:helmholtz} after application of padding, data
%  alignment, and \emph{licm}, for an AVX architecture. In this
%  example, sub-expressions invariant to \texttt{j} are identical to
%  those invariant to \texttt{k}, so they can be precomputed once in
%  the $r$ loop.}
%\label{code:helmholtz-licm}
%\end{algorithm}

%\begin{figure}[b]
%\begin{center}
%\includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
%\caption{High-level view of Firedrake. COFFEE is at the core,
%  receiving ASTs from a modified version of the FEniCS Form compiler
%  and producing optimized C code kernels.}
%\label{fig:coffee-pipeline}
%\end{center}
%\end{figure}

>>>>

\begin{figure}
\tiny
\lstinputlisting{listings/simple.code}
\caption{Original code}\label{fig:original-code}
\end{figure}

@@

\begin{figure}
\tiny
\lstinputlisting{listings/factorized.code}
\caption{Factorized code}\label{fig:factorized-code}
\end{figure}

@@

\begin{figure}
\tiny
\lstinputlisting{listings/toexpand.code}
\caption{Expandable code}\label{fig:toexpand-code}
\end{figure}

@@

%TODO: each level in sigma is then taken and wrapped in a loop

@@

\subsection{Avoiding Iteration on Zero-blocks by Symbolic Execution}
\label{sec:zeros}
The second task of the ER consists of skipping arithmetic operations over blocks of zero-valued entries in basis function and derivatives of basis functions arrays. Zero columns in such arrays arise, for example, when taking derivatives on a reference element and when employing mixed elements. In~\cite{quadrature1}, a code transformation to eliminate floating point operations on zero columns was introduced; the technique was implemented in FEniCS. We will evaluate our approach and compare it to this pioneering work in Section~\ref{sec:perf-results}. We propose a strategy in which we avoid the use of indirection arrays in the generated code (e.g. \texttt{A[B[i]]}, where \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in \texttt{A}), which otherwise would break the optimizations applicable at the Code Specializer level, especially SIMD vectorization.

\begin{figure}[t]
\tiny
\centering     
\subfigure[Factorized, showing zeros, code]{\label{fig:withzeros-code}\lstinputlisting{listings/withzeros.code}}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subfigure[Factorized, skipping zeros, code]{\label{fig:withzeros-skipped-code}\lstinputlisting{listings/skipzeros.code}}
\caption{Without and with skipping zeros}\label{fig:skip-code}
\end{figure}

Consider Figure~\ref{fig:withzeros-code}, which is an enriched version of the transformed Burgers excerpt in Figure~\ref{fig:factorized-code}. The code is instantiated for the specific case of polynomial order 1, Lagrange elements on a 2D mesh. The array \texttt{D} represents a tabulated derivative of a basis function at the various quadrature points. We notice the presence of 4 zero-columns. Any multiplications or additions along these columns could (should) be skipped in order to minimize irrelevant floating point operations. One solution, as anticipated, is not to generate the zero columns, i.e. to generate a dense 6$\times$2 array, to reduce the size of the iteration space over test and trial functions (from 6 to 2), and to use an indirection array (e.g. $ind = \lbrace 3, 5\rbrace$) to access the right entries in the element tensor $A$. As explained, this prevents, among the various optimizations, effective code vectorizability, because memory loads and store would eventually reference non-contiguous locations. 
%on the other hand, reshrinki tutto su un array solo...

Our approach is based on domain knowledge and symbolic execution. We discern the origin of zero-columns: for example, those due to taking derivatives on the reference element from those inherent to using mixed (vector) elements. In the running Burgers example, the use of vector function spaces require the generation of a zero-block (columns 0, 1, 2 in the array \texttt{D}) to preserve the correctness of the element matrix evaluation while iterating along the space of test and trial functions. The two key observations are that 1) the number of zero-columns caused by mixed elements is much larger then that due to derivatives, and 2) such columns are contiguous in memory. Based on this, we decide to skip any computation involving zero-columns induced by mixed elements, whereas we still iterate over any other zero-columns. Consequently, the only modifications to the code that are eventually needed are adjusting loop bounds and introducing offsets when accessing the element matrix, as shown in Figure~\ref{fig:withzeros-skipped-code}. In general, multiple iteration spaces over test and trial functions (i.e. \texttt{jk} loops), characterized by their own loop bounds, may be needed, increasing loop overhead and decreasing data locality for the element matrix.

The implementation is based on symbolic execution. The ER expects information from the higher layer about where the zero-blocks are located, in each tabulated basis function. This could come in the form of code annotation if the input to COFFEE were provided as a C string (as shown in Figure~\ref{fig:withzeros-skipped-code}) or by suitably decorating basis functions nodes in the abstract syntax tree representing the local assembly routine. Then, each statement in the local assembly loop nest is symbolically executed. For example, consider the statement \texttt{T2[r] = ((d*D[i][k])+(e*E[i][k]));} in Figure~\ref{fig:withzeros-skipped-code}. Array \texttt{D} has non-zeros in positions $NZ_D = [3,5]$; we also assume \texttt{E} has non-zeros in positions $NZ_E = [0,2]$. Multiplications by scalar do not affect the propagation of zero columns. On the other hand, when summing the two operands \texttt{d*D[i][k]} and \texttt{e*E[i][k]}, we record that the target identifier \texttt{T2} will have non-zero columns in positions $NZ_D \cup NZ_E = [0-5]$. Exploiting the $NZ$ information associated with each identifier, the assembly expression is split into several sets of sub-expressions, each set characterized by accesses to the same range of non-zero columns and, therefore, iterating over a suitably-sized iteration space. In our example, there is just one set, which contains two sub-expressions, enclosed in a doubly nested loop of size 3$\times$3.

\section{Code Specialization}
\label{sec:code-spec}
The Code Specializer is the second software component of COFFEE. It provides a range of code transformations tailored for optimizing instruction-level parallelism and register locality. As summarized in Section~\ref{sec:summary-opts} and described in~\cite{Luporini}, padding and data alignment, expression splitting, and outer-product vectorization, which is a domain-aware implementation of vector-register tiling, are examples of optimizations that the Code Specializer is capable of applying. In this paper, we enrich this set of transformations as detailed in the following sections.

\subsection{Vector-expansion of Invariant Terms}
The Expression Rewriter's loop-invariant code motion hoists invariant sub-expressions in the body of the outermost loop along which they iterate. If such a sub-expression computes a value depending on the outermost loop only, then the only way of vectorizing it - if we exclude superword level parallelism~\cite{SLP}, which is in general not applicable to our kernels - is by vector-expansion of terms.

\begin{figure}[t]
\tiny
\centering     
\subfigure[To vector expand code]{\label{fig:tovectexpand-code}\lstinputlisting{listings/tovectexpand.code}}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subfigure[Vector-expanded code]{\label{fig:vectexpanded-code}\lstinputlisting{listings/vectexpanded.code}}
\caption{Before and after vector expansion}\label{fig:skip-code}
\end{figure}

An example is given in Figure~\ref{fig:tovectexpand-code}: \texttt{F1, F2, F3, T1} are all scalar quantities, and their computation will not be auto-vectorized (as explained in Section~\ref{sec:perf-results}, we experimented with the Intel and GNU compilers). The argument that the cost of evaluating such expressions is not relevant because their computational cost is $O(I)$, while the evaluation of the element matrix has a cost of $O(I \cdot T^2)$, is too superficial, in general. There are notable cases, for instance problems based on hyperlasticity, for which the Expression Rewriter produces a significantly-large number of invariant scalar terms, which dramatically impact the execution time.

We have implemented a procedure that vector-expands scalar expressions, while lifting them outside of the loop nest as shown in Figure~\ref{fig:vectexpanded-code}. The dependency graph is queried to assess the legality of the transformation. At the price of extra memory, vector-expanded terms can now be vectorized by the compiler.

% Expose vectorizability of the outermost loop 
% Render the loop nest perfect

@@

% can i just add a rewrite rule to expose these stuff?

\subsection{Model-driven Autotuning}
%\subsection{Standard Compiler Transformations}
